---
title: "Manage .data.yaml from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Manage .data.yaml from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
live <- identical(Sys.getenv("DATUR_BUILD_LIVE_VIGNETTES"), "true")
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = live)

if (live) {
  library(datur)
  stopifnot(datum_available())
  project <- tempfile("datur-data-yaml-")
  dir.create(file.path(project, "upstream"), recursive = TRUE)
  writeLines(
    sprintf("date,rate\n%s,1", Sys.Date()),
    file.path(project, "upstream", "rates.csv")
  )
  writeLines("id,value\n1,retained", file.path(project, "upstream", "retained.csv"))
}
```

Current `datum` releases expose the exact configuration schema in the binary,
together with the required and optional fields for every source handler
included in the build. `datur` uses that live metadata to make `.data.yaml`
edits safer: an edit is validated before the original file is replaced, and a
failed validation leaves it untouched.

The executable is still the source of truth. These helpers verify that the
installed release supports configuration metadata, even though checking
existing projects also works with older supported releases.

## Inspect the installed configuration contract

`datum_schema()` parses the JSON Schema returned by `datum schema`:

```{r schema}
schema <- datum_schema()
schema$title
schema$properties$datasets$items$properties$policy$enum
```

`datum_types()` reports the source handlers available in this particular
build. That distinction matters because optional build tags can change the
available handlers.

```{r types}
names(datum_types())
datum_types(c("http", "file"))
```

## Keep secrets outside `.data.yaml` (datum 1.5.0+)

Starting with `datum` 1.5.0, any YAML string value can contain a `${NAME}`
environment reference. `datum` expands it after parsing the YAML and before
validating or using the configuration. The placeholder can therefore be
written with `datur` just like any other source or target value:

```{r secret-source, eval=FALSE}
private_api <- datum_source(
  "http",
  url = "https://api.example.com/export?token=${API_TOKEN}"
)

datum_dataset_add(
  id = "private_export",
  desc = "Authenticated API export",
  target = "${DATA_DIR}/private.csv",
  source = private_api,
  policy = "update"
)
```

Supply the variables through the environment inherited by `datum`. For
example, they may be injected by a CI secret store or set in the R process
before the call:

```{r secret-environment, eval=FALSE}
Sys.setenv(
  API_TOKEN = "value-supplied-at-runtime",
  DATA_DIR = "data"
)
datum_check(quiet = TRUE)
```

The expansion rules are deliberately strict:

- variable names must match `[A-Za-z_][A-Za-z0-9_]*`;
- an unset variable is a configuration error, while a variable explicitly set
  to an empty string is allowed;
- only `${NAME}` is expanded, so plain `$NAME` remains available to command
  sources;
- `$${NAME}` produces the literal text `${NAME}`; and
- substitution applies to YAML values, not mapping keys.

Quote a placeholder that replaces a whole value when the expanded value must
remain a string. Placeholders remain in `.data.yaml`, and expanded values are
not written to `.data.lock.yaml`. This prevents committing the resolved secret,
but it is not an output-redaction boundary: downstream URLs, commands, and
errors may still expose it. Prefer purpose-built authentication variables such
as the Git handler's `GIT_TOKEN` when available, and do not print secrets from
command sources.

`datur`'s configuration helpers validate the placeholder-bearing document
before writing it. Retain any literal prefix required by the schema—for
example, keep `https://` outside the placeholder in an HTTP URL. `datum` then
validates the expanded value again at runtime.

## Construct a source

`datum_source()` checks field names and required values against
`datum_types()`. It returns an ordinary named list, so you can inspect, store,
or modify it before adding a dataset.

```{r sources}
primary <- datum_source(
  "file",
  path = "upstream/rates.csv"
)

fallback <- datum_source(
  "file",
  path = "offline-cache/rates.csv"
)
```

Unknown fields, missing required fields, unavailable source types, and empty
values fail before `.data.yaml` is touched.

## Add datasets

Use `source` for one source or `sources` for ordered fallback. If the config
does not exist, `datum_dataset_add()` starts a version 1 document. Paths are
interpreted by datum relative to the project working directory, so pass the
same `wd` you use for checks when working outside the current R directory.

```{r add}
datum_dataset_add(
  id = "exchange_rates",
  desc = "Daily exchange rates",
  target = "data/rates.csv",
  sources = list(primary, fallback),
  policy = "update",
  wd = project
)

datum_dataset_add(
  id = "retained_data",
  desc = "A second local dataset",
  target = "data/retained.csv",
  source = datum_source("file", path = "upstream/retained.csv"),
  policy = "update",
  wd = project
)
```

The full resulting document—not only the new dataset—is checked against the
installed schema and the available source types. Existing defaults and other
datasets are retained.

## Update datasets

Only arguments you supply are changed. Supplying `source` replaces an existing
`sources` list, and supplying `sources` replaces `source`. Use `policy = NULL`
to remove a dataset-level override and fall back to the config default.

```{r update}
datum_dataset_update(
  "exchange_rates",
  desc = "Rates with an offline fallback",
  policy = NULL,
  wd = project
)

datum_dataset_update(
  "exchange_rates",
  source = primary,
  wd = project
)
```

Dataset identifiers are immutable in this API. Remove and re-add a definition
when an identifier itself must change, then decide how its old lock entry
should be handled.

## Audit and remove datasets

`datum_audit()` combines config and lockfile state without contacting any data
source. Its records distinguish configured and pinned datasets from pending,
deleted, and orphaned entries.

```{r audit}
audit <- datum_audit(wd = project)
as.data.frame(audit)

subset(as.data.frame(audit), status != "ok")
```

`datum_dataset_remove()` removes a definition from `.data.yaml`. In an
interactive R session it asks whether the tracked local files should also be
deleted. In a script, state the choice explicitly:

```{r remove}
# Keep the downloaded file.
datum_dataset_remove("exchange_rates", delete = FALSE, wd = project)
```

Pass `delete = TRUE` instead to run `datum delete` before removing the YAML
definition.

Deletion happens first because datum needs the still-present definition to
locate the target. It marks the lock entry deleted; after removing the config
definition, `datum_audit()` can still show that historical entry. Use datum's
low-level `unlock` command through `datum_run()` if you deliberately want to
forget it completely.

To delete local files while keeping the dataset definition, call
`datum_delete()` directly. It asks for confirmation in an interactive session;
automation must opt in with `yes = TRUE`.

```{r prepare-delete, include=FALSE}
writeLines("id,value\n1,cached", file.path(project, "upstream", "cached.csv"))
datum_dataset_add(
  "cached_data", "Cached local data", "data/cached.csv",
  datum_source("file", path = "upstream/cached.csv"),
  policy = "update", wd = project
)
datum_check(wd = project, quiet = TRUE)
```

```{r delete}
datum_delete("cached_data", yes = TRUE, wd = project)
```

Datum then skips deleted datasets until its `undelete` command is run. The
`.data.yaml` file itself is never changed by `datum_delete()`.
