---
title: "Manage .data.yaml from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Manage .data.yaml from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Datum 1.4.0 exposes the exact configuration schema in its binary, together
with the required and optional fields for every source handler included in the
build. `datur` uses that live metadata to make `.data.yaml` edits safer: an edit
is validated before the original file is replaced, and a failed validation
leaves it untouched.

The executable is still the source of truth. These helpers require datum 1.4.0
or newer even though checking existing projects works with older supported
1.x releases.

## Inspect the installed configuration contract

`datum_schema()` parses the JSON Schema returned by `datum schema`:

```{r schema, eval=FALSE}
schema <- datum_schema()
schema$title
schema$properties$datasets$items$properties$policy$enum
```

`datum_types()` reports the source handlers available in this particular
build. That distinction matters because optional build tags can change the
available handlers.

```{r types, eval=FALSE}
names(datum_types())
datum_types(c("http", "file"))
```

## Construct a source

`datum_source()` checks field names and required values against
`datum_types()`. It returns an ordinary named list, so you can inspect, store,
or modify it before adding a dataset.

```{r sources, eval=FALSE}
primary <- datum_source(
  "http",
  url = "https://example.test/rates.csv"
)

fallback <- datum_source(
  "file",
  path = "offline-cache/rates.csv"
)
```

Unknown fields, missing required fields, unavailable source types, and empty
values fail before `.data.yaml` is touched.

## Add datasets

Use `source` for one source or `sources` for ordered fallback. If the config
does not exist, `datum_dataset_add()` starts a version 1 document. Paths are
interpreted by datum relative to the project working directory, so pass the
same `wd` you use for checks when working outside the current R directory.

```{r add, eval=FALSE}
project <- "/analysis/exchange-model"

datum_dataset_add(
  id = "exchange_rates",
  desc = "Daily exchange rates",
  target = "data/rates.csv",
  sources = list(primary, fallback),
  policy = "update",
  wd = project
)
```

The full resulting document—not only the new dataset—is checked against the
installed schema and the available source types. Existing defaults and other
datasets are retained.

## Update datasets

Only arguments you supply are changed. Supplying `source` replaces an existing
`sources` list, and supplying `sources` replaces `source`. Use `policy = NULL`
to remove a dataset-level override and fall back to the config default.

```{r update, eval=FALSE}
datum_dataset_update(
  "exchange_rates",
  desc = "Rates with an offline fallback",
  policy = NULL,
  wd = project
)

datum_dataset_update(
  "exchange_rates",
  source = primary,
  wd = project
)
```

Dataset identifiers are immutable in this API. Remove and re-add a definition
when an identifier itself must change, then decide how its old lock entry
should be handled.

## Audit and remove datasets

`datum_audit()` combines config and lockfile state without contacting any data
source. Its records distinguish configured and pinned datasets from pending,
deleted, and orphaned entries.

```{r audit, eval=FALSE}
audit <- datum_audit(wd = project)
as.data.frame(audit)

subset(as.data.frame(audit), status != "ok")
```

`datum_dataset_remove()` removes a definition from `.data.yaml`. In an
interactive R session it asks whether the tracked local files should also be
deleted. In a script, state the choice explicitly:

```{r remove, eval=FALSE}
# Keep the downloaded file.
datum_dataset_remove("exchange_rates", delete = FALSE, wd = project)

# First run datum delete, then remove the YAML definition.
datum_dataset_remove("exchange_rates", delete = TRUE, wd = project)
```

Deletion happens first because datum needs the still-present definition to
locate the target. It marks the lock entry deleted; after removing the config
definition, `datum_audit()` can still show that historical entry. Use datum's
low-level `unlock` command through `datum_run()` if you deliberately want to
forget it completely.

To delete local files while keeping the dataset definition, call
`datum_delete()` directly. It asks for confirmation in an interactive session;
automation must opt in with `yes = TRUE`.

```{r delete, eval=FALSE}
datum_delete("exchange_rates", wd = project)
datum_delete(c("large_one", "large_two"), yes = TRUE, wd = project)
```

Datum then skips deleted datasets until its `undelete` command is run. The
`.data.yaml` file itself is never changed by `datum_delete()`.
