---
title: "A Complete datum Workflow from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Complete datum Workflow from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`datur` is deliberately a thin integration layer. `datum` owns source access,
fingerprinting, policies, and lockfile updates; `datur` owns safe process
execution, protocol validation, typed conditions, and convenient R objects.

This vignette is self-contained during package checks: commands requiring the
external executable are not evaluated while knitting. Output labelled **datum
1.3.0 output** was captured verbatim from a release-tag build using local file
sources. No network requests or credentials were involved.

## 1. Discover and validate the executable

```{r discovery, eval=FALSE}
library(datur)

datum_available()
#> [1] TRUE

datum_path()
#> [1] "/usr/local/bin/datum"

datum_version()
#> [1] '1.3.0'
```

Discovery precedence is explicit and deterministic:

1. `executable =` on the function call;
2. `options(datur.datum_path = ...)`;
3. `DATUM_PATH`;
4. `Sys.which("datum")`.

Automatic `PATH` discovery and successful version parsing are cached. Use
`datum_path(refresh = TRUE)` or `datum_version(refresh = TRUE)` after replacing
the executable in a long-running R session.

## 2. Configure a feature-rich datum project

The following configuration demonstrates strict checks, automatic updates,
monitoring, and source fallback. It is a datum file, not an R-specific format.

```yaml
version: 1

defaults:
  policy: fail
  algo: sha256

datasets:
  - id: customer_reference
    desc: Versioned customer reference data
    source:
      type: file
      path: upstream/customers.csv
    target: data/customers.csv
    policy: fail

  - id: exchange_rates
    desc: Rates refreshed automatically
    sources:
      - type: http
        url: https://primary.example.test/rates.csv
      - type: file
        path: offline-cache/rates.csv
    target: data/rates.csv
    policy: update

  - id: model_license
    desc: Monitor a file from a Git repository
    source:
      type: git
      url: https://github.com/example/model.git
      ref: main
      path: LICENSE
    target: vendor/model-LICENSE
    policy: log
```

Relative config, lock, source, and target paths are resolved from the child
working directory. Set that directory without touching R's global working
directory:

```{r isolated-working-directory, eval=FALSE}
project <- "/analysis/customer-model"

result <- datum_check(
  config = ".data.yaml",
  lock = ".data.lock.yaml",
  wd = project,
  timeout = 120,
  concurrency = 1,
  quiet = TRUE
)

stopifnot(identical(getwd(), original_r_working_directory))
```

Concurrency defaults to one because datum 1.x documents a race when datasets
sharing a target directory are fetched concurrently.

## 3. Understand the actual JSON boundary

`datur` invokes flags before the subcommand:

```text
datum --json --no-color --timeout 120s --concurrency 1 check
```

For one unchanged local-file target, datum 1.3.0 emitted:

```json
{
  "results": [
    {
      "id": "local_config",
      "status": "ok",
      "lock_fingerprint": "sha256:ece015bd6bc8299cc3085ffd5848a841a9c7faf2912ec05176f10bda2f81e2af",
      "remote_fingerprint": "sha256:ece015bd6bc8299cc3085ffd5848a841a9c7faf2912ec05176f10bda2f81e2af"
    }
  ]
}
```

The process exited 0. In R, the normalized object prints concisely:

```{r print-unchanged, eval=FALSE}
print(result)
#> datum check: no changes across 1 target (8ms)

result$changed
#> [1] FALSE

result$status
#> [1] "unchanged"
```

The complete machine document remains available as `result$raw`, while
documented normalized fields remain stable if datum adds JSON fields later.

## 4. Treat changed data as data

After replacing the recorded fingerprint with a synthetic older value, a
`fail`-policy check produced this **datum 1.3.0 output** and exit status 1:

```json
{
  "results": [
    {
      "id": "local_config",
      "status": "fail",
      "lock_fingerprint": "sha256:0000000000000000000000000000000000000000000000000000000000000000",
      "remote_fingerprint": "sha256:ece015bd6bc8299cc3085ffd5848a841a9c7faf2912ec05176f10bda2f81e2af"
    }
  ]
}
```

That exit code is not an R error because the JSON status proves this is a
detected change:

```{r inspect-change, eval=FALSE}
result <- datum_check(wd = project, quiet = TRUE)
print(result)
#> datum check: 1 change across 1 target (8ms)
#> fail local_config

result$changed
#> [1] TRUE

result$process$status
#> [1] 1

as.data.frame(result)[, c("id", "status", "lock_fingerprint", "remote_fingerprint")]
#>             id status                                                  lock_fingerprint
#> 1 local_config   fail sha256:0000000000000000000000000000000000000000000000000000000000000000
#>                                                  remote_fingerprint
#> 1 sha256:ece015bd6bc8299cc3085ffd5848a841a9c7faf2912ec05176f10bda2f81e2af
```

The same change under `log` has status `stale` and process status 0. Under
`update`, status `updated` and process status 0 mean datum fetched the target and
updated its lock record. All three statuses set `result$changed` to `TRUE`.

## 5. Build an incremental pipeline

```{r pipeline, eval=FALSE}
result <- datum_check(wd = project, quiet = TRUE)
records <- as.data.frame(result)

changed <- records$status %in% c("updated", "stale", "fail")
changed_ids <- records$id[changed]

if ("customer_reference" %in% changed_ids) {
  source("R/rebuild-customer-features.R")
}

if ("exchange_rates" %in% changed_ids) {
  rmarkdown::render("reports/exposure.Rmd")
}

saveRDS(
  list(
    checked_at = result$checked_at,
    datum_version = as.character(result$datum_version),
    records = records
  ),
  "artifacts/data-check.rds"
)
```

Useful stable fields include:

- `changed`: one non-missing logical value;
- `status`: `unchanged`, `changed`, `partial_failure`, or `error`;
- `summary`: named counts for total, changed, unchanged, failed, and skipped;
- `records`: one base data-frame row per datum dataset;
- `checked_at` and numeric `duration` in seconds;
- `datum_version` and implicit `schema_version` 1;
- `process`: redacted command metadata and captured streams.

## 6. Preserve partial results when a source fails

Datum continues across datasets. A local test with one available and one missing
source emitted the following document and exited 1:

```json
{
  "results": [
    {
      "id": "available",
      "status": "ok",
      "lock_fingerprint": "sha256:ece015bd6bc8299cc3085ffd5848a841a9c7faf2912ec05176f10bda2f81e2af",
      "remote_fingerprint": "sha256:ece015bd6bc8299cc3085ffd5848a841a9c7faf2912ec05176f10bda2f81e2af"
    },
    {
      "id": "unavailable",
      "status": "error",
      "message": "fingerprint: stat missing-source.json: no such file or directory"
    }
  ]
}
```

`datur` raises `datur_cli_error`, but attaches the normalized partial result:

```{r partial-failure, eval=FALSE}
result <- tryCatch(
  datum_check(wd = project, quiet = TRUE),
  datur_cli_error = function(error) {
    if (is.null(error$result)) stop(error)
    message(conditionMessage(error))
    error$result
  }
)

result$status
#> [1] "partial_failure"

subset(as.data.frame(result), status == "error", c("id", "message"))
#>            id                                                   message
#> 2 unavailable fingerprint: stat missing-source.json: no such file or directory
```

This supports a deliberate policy: use successful records, alert on failures,
and decide whether the surrounding pipeline should stop.

## 7. Respond to typed conditions

All package errors inherit from `datur_error` and a specific subclass:

```{r conditions, eval=FALSE}
tryCatch(
  datum_check(wd = project, timeout = 30, quiet = TRUE),
  datur_not_found = function(error) {
    message("Install datum or configure DATUM_PATH")
  },
  datur_timeout = function(error) {
    message("Timed out after ", error$timeout, " seconds")
    print(error$partial_result)
  },
  datur_protocol_error = function(error) {
    message("Unsupported or malformed datum JSON")
    print(error$process)
  },
  datur_cli_error = function(error) {
    message("datum operational failure: ", conditionMessage(error))
  }
)
```

A change alone never raises a warning or error. Operational source failures,
lock write failures, malformed JSON, incompatible versions, and R input errors
remain distinguishable.

## 8. Use the low-level escape hatch intentionally

```{r low-level, eval=FALSE}
process <- datum_run(
  args = c("--json", "audit"),
  wd = project,
  timeout = 30,
  error_on_status = FALSE
)

print(process)
#> datum: status 0 (18ms)

audit <- jsonlite::fromJSON(process$stdout)
audit$entries
```

`datum_run()` sends each `args` element directly to the executable. It does not
construct a shell command. Named `env` values augment the inherited environment,
and character or raw `stdin` is supplied through a private temporary file.

## 9. Know the side effects and security boundary

Even though its name is `check`, datum always rewrites lockfile timestamps. An
`update` policy may fetch and replace targets, and source handlers may use the
network or datum's git cache. Run checks in a project where those effects are
expected.

Do not put credentials directly in URLs or command output. `datur` redacts
sensitive argument values, sensitive environment values, and URL user info from
public process results, but the safest design is still to keep secrets in the
credential mechanisms datum documents.

For the exact reviewed boundary, including statuses and exit-code ambiguities,
see `docs/datum-cli-contract.md` in the source repository.
