---
title: "A Complete datum Workflow from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Complete datum Workflow from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
live <- identical(Sys.getenv("DATUR_BUILD_LIVE_VIGNETTES"), "true")
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = live)

if (live) {
  library(datur)
  stopifnot(datum_available())

  project <- tempfile("datur-vignette-")
  dir.create(file.path(project, "upstream"), recursive = TRUE)
  writeLines("name,value\nalpha,1", file.path(project, "upstream", "customers.csv"))

  local_file <- datum_source("file", path = "upstream/customers.csv")
  datum_dataset_add(
    id = "customer_reference",
    desc = "Versioned customer reference data",
    target = "data/customers.csv",
    source = local_file,
    policy = "update",
    wd = project
  )
  datum_check(wd = project, quiet = TRUE)
  datum_dataset_update("customer_reference", policy = "fail", wd = project)
}
```

`datur` is deliberately a thin integration layer. `datum` owns source access,
fingerprinting, policies, and lockfile updates; `datur` owns safe process
execution, protocol validation, typed conditions, and convenient R objects.

The pkgdown version of this vignette is built against the latest published
`datum` release. Its output is produced during the site build from disposable
local files; no output transcript is stored in the vignette source.

## 1. Install or discover the executable

Installing `datum` through `datur` is always explicit. The default resolves the
latest GitHub release and chooses the archive matching the current operating
system and CPU architecture:

```{r download, eval=FALSE}
download <- datum_download()
if (download$downloaded) {
  options(datur.datum_path = download$path)
}
```

For a reproducible toolchain, reuse the resolved release and pin the
destination:

```{r pinned-download, eval=FALSE}
datum_download(
  version = download$version,
  destination = "~/bin/datum",
  overwrite = TRUE
)
```

The release archive is checked against `checksums.txt` before its executable is
installed. If GitHub exceeds the requested timeout, no partial binary is
installed; the returned object's `url` and `asset` fields identify the manual
download instead.

Alternatively, install `datum` separately and let `datur` discover it:

```{r discovery}
datum_available()
datum_path()
datum_version()
```

Discovery precedence is explicit and deterministic:

1. `executable =` on the function call;
2. `options(datur.datum_path = ...)`;
3. `DATUM_PATH`;
4. `Sys.which("datum")`.

Automatic `PATH` discovery and successful version parsing are cached. Use
`datum_path(refresh = TRUE)` or `datum_version(refresh = TRUE)` after replacing
the executable in a long-running R session.

## 2. Configure a datum project

This site build created a disposable project with a local-file source, fetched
it once, and changed its policy to `fail`. Here is the configuration written by
`datur` using the schema embedded in the installed executable:

```{r configuration}
cat(readLines(file.path(project, ".data.yaml")), sep = "\n")
```

Relative config, lock, source, and target paths are resolved from the child
working directory. Set that directory without touching R's global working
directory:

```{r isolated-working-directory}
original_r_working_directory <- getwd()
result <- datum_check(
  config = ".data.yaml",
  lock = ".data.lock.yaml",
  wd = project,
  timeout = 120,
  concurrency = 1,
  quiet = TRUE
)

stopifnot(identical(getwd(), original_r_working_directory))
```

Concurrency defaults to one because datum documents a race when datasets
sharing a target directory are fetched concurrently.

## 3. Understand the actual JSON boundary

`datur` invokes flags before the subcommand:

```text
datum --json --no-color --timeout 120s --concurrency 1 check
```

The process response is retained rather than copied into the documentation:

```{r json-unchanged}
cat(jsonlite::prettify(result$process$stdout))
```

The normalized object prints concisely:

```{r print-unchanged}
print(result)
result$changed
result$status
```

The complete machine document remains available as `result$raw`, while
documented normalized fields remain stable if datum adds JSON fields later.

## 4. Treat changed data as data

Change the local source after it has been fetched to exercise the `fail`
policy with the installed `datum` version.

That exit code is not an R error because the JSON status proves this is a
detected change:

```{r inspect-change}
writeLines("name,value\nalpha,2", file.path(project, "upstream", "customers.csv"))
result <- datum_check(wd = project, quiet = TRUE)
print(result)
result$changed
result$process$status
as.data.frame(result)[, c("id", "status", "lock_fingerprint", "remote_fingerprint")]
```

The same change under `log` has status `stale` and process status 0. Under
`update`, status `updated` and process status 0 mean datum fetched the target and
updated its lock record. All three statuses set `result$changed` to `TRUE`.

## 5. Build an incremental pipeline

```{r pipeline, eval=FALSE}
result <- datum_check(wd = project, quiet = TRUE)
records <- as.data.frame(result)

changed <- records$status %in% c("updated", "stale", "fail")
changed_ids <- records$id[changed]

if ("customer_reference" %in% changed_ids) {
  source("R/rebuild-customer-features.R")
}

if ("exchange_rates" %in% changed_ids) {
  rmarkdown::render("reports/exposure.Rmd")
}

saveRDS(
  list(
    checked_at = result$checked_at,
    datum_version = as.character(result$datum_version),
    records = records
  ),
  "artifacts/data-check.rds"
)
```

Useful stable fields include:

- `changed`: one non-missing logical value;
- `status`: `unchanged`, `changed`, `partial_failure`, or `error`;
- `summary`: named counts for total, changed, unchanged, failed, and skipped;
- `records`: one base data-frame row per datum dataset;
- `checked_at` and numeric `duration` in seconds;
- `datum_version` and implicit `schema_version` 1;
- `process`: redacted command metadata and captured streams.

## 6. Preserve partial results when a source fails

Datum continues across datasets. Add a missing local source to demonstrate that
`datur` raises `datur_cli_error` but attaches the normalized partial result:

```{r prepare-partial, include=FALSE}
datum_dataset_update("customer_reference", policy = "update", wd = project)
datum_check(wd = project, quiet = TRUE)
datum_dataset_update("customer_reference", policy = "fail", wd = project)
datum_dataset_add(
  id = "unavailable",
  desc = "A deliberately missing local source",
  target = "data/unavailable.json",
  source = datum_source("file", path = "upstream/missing-source.json"),
  policy = "fail",
  wd = project
)
```

```{r partial-failure}
result <- tryCatch(
  datum_check(wd = project, quiet = TRUE),
  datur_cli_error = function(error) {
    if (is.null(error$result)) stop(error)
    message(conditionMessage(error))
    error$result
  }
)

result$status
subset(as.data.frame(result), status == "error", c("id", "message"))
```

This supports a deliberate policy: use successful records, alert on failures,
and decide whether the surrounding pipeline should stop.

## 7. Respond to typed conditions

All package errors inherit from `datur_error` and a specific subclass:

```{r conditions, eval=FALSE}
tryCatch(
  datum_check(wd = project, timeout = 30, quiet = TRUE),
  datur_not_found = function(error) {
    message("Install datum or configure DATUM_PATH")
  },
  datur_timeout = function(error) {
    message("Timed out after ", error$timeout, " seconds")
    print(error$partial_result)
  },
  datur_protocol_error = function(error) {
    message("Unsupported or malformed datum JSON")
    print(error$process)
  },
  datur_cli_error = function(error) {
    message("datum operational failure: ", conditionMessage(error))
  }
)
```

A change alone never raises a warning or error. Operational source failures,
lock write failures, malformed JSON, incompatible versions, and R input errors
remain distinguishable.

## 8. Use the low-level escape hatch intentionally

```{r low-level}
process <- datum_run(
  args = "--version",
  wd = project,
  timeout = 30,
  error_on_status = FALSE
)

print(process)
process$stdout
```

`datum_run()` sends each `args` element directly to the executable. It does not
construct a shell command. Named `env` values augment the inherited environment,
and character or raw `stdin` is supplied through a private temporary file.

## 9. Know the side effects and security boundary

Even though its name is `check`, datum always rewrites lockfile timestamps. An
`update` policy may fetch and replace targets, and source handlers may use the
network or datum's git cache. Run checks in a project where those effects are
expected.

Do not put credentials directly in URLs or command output. `datur` redacts
sensitive argument values, sensitive environment values, and URL user info from
public process results, but the safest design is still to keep secrets in the
credential mechanisms datum documents.

For the exact reviewed boundary, including statuses and exit-code ambiguities,
see `docs/datum-cli-contract.md` in the source repository.
